Random Generation of Rich \\ Abstract Data Type Values

### 2019-01-31 paper/synthesis.tex
Pass over predictions

Essentially, a branching process is a special kind of Markov process that models
the evolution of a population of \emph{individuals of different kinds} across
discrete time steps known as \emph{generations}.
%
Each kind of individual is expected to produce an average number of offspring of
(possibly) different kinds from one generation to the next one.
%
Mista el at. \cite{DBLP:conf/haskell/MistaRH18} show that branching processes
can be adapted to predict the generation of ADT values by simply considering
each data constructor as an individual on its own.
%
In fact, any ADT value can be seen as a tree where each node represents a root
data constructor and has its sub-expressions as sub-trees---hence note the
similarity with family trees.
%
In this light, each tree level of a random value can be seen as a generation of
individuals in this model.

@@

Vector $E[G_0]$ represents the initial distribuition of constructors that our
generator produces, which simply consists of the generation probability of each
one.
%
The interesting aspect of the prediction mechanism is encoded in the matrix $M$,
known as a the \emph{mean matrix} of this stochastic process.
%
$M$ is a squared matrix with as rows and columns as different data constructors
involved in the generation process.
%
Each element $M_{i,j}$ of this matrix encodes the average number of data
constructors of kind $j$ that gets generated in a given level, provided that we
generated a constructor of kind $i$ at the previous level.
%
In this sense, this matrix encodes the ``branching'' behavior of our random
generation from one generation to the next one.
%
Each element of the matrix can be automatically calculated by exploiting ADT
definitions \cite{DBLP:conf/haskell/MistaRH18}
%
% ---specially the recursive fields---
%
as well as the individual probability of generating each constructor.
%
For instance, the average number of |Text| data constructors that we will
generate provided that we generated a |Join| data constructor on the previous
level is calculated as follows:
%
$$ M_{Join, Text} = 2\cdot p_{Text} $$
%
where $2$ is the number of holes present when generating a partial ADT value
|Join| (i.e., |Join hole hole|) and $p_{Text}$ is the probability of
individually generating the constructor |Text|.
%
This pattern can be used to build the rest of the mean matrix as well.

@@

The new cells are filled as before: we need to consider the amount of holes when
generating partial pattern matching and function calls as well as their
individual probabilities.
%
% Fortunately, it is possible to follow the same pattern as we did on our original
% work, this time considering the amount of recursive pattern variables if it the
% case of a pattern matching value being a father individual, and the amount of
% recursive arguments if we deal with a abstract interface function call as a
% father individual.
%
For instance, if we consider $P_j$ as the second pattern of simplify and $F_1$
as function |div|, then the marked cell above has the value $2\cdot p_{div}$,
i.e., the amount of holes in a partially generated pattern |(Join (Join (Text s)
hole) hole)|, where |s| is some random string, times the probability to generate
a call to function |div|.
%
The rest of this matrix can be computed analogously.

@@

As another contribution, we found that the whole prediction process can be
factored in terms of two vectors $\beta$ and $\mathcal{P}$, such that $\beta$
represents the number of recursive sub-terms of each construction that we
generate (i.e., data constructors, pattern matching values or function calls),
whereas $\mathcal{P}$ simply represents the probability of generating that
partial ADT value.% (this probability can be easily calculated based on the
% generation frequencies used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewriten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P})^{n}
\end{align*}
%
The vector $\beta$ can be obtained by analyzing the shape (namely holes) of each
source of structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as shown in Fig. \ref{fig:vectors} (where dashed lines indicate the
composition of the diffent sub-vectors that are calculated separately).

@@

It is possible to provide an overall prediction of the expected number of
constructors when restricting the generation process to only bare data
constructors and pattern matching values.
%
To achieve that, we should stop considering pattern matching values as atomic
values and start seeing them as compositions of several data constructors.
%
In that manner, it is possible to obtain the expected \emph{total} number of
generated data constructors that our generators will produce---regardless if they
are generated on their own, or as part of a pattern matching value.
%
We note this number as $E^\downarrow\![\_]$ and, to calculate it, we only need
to add the expected number of bare constructors that are included within each
pattern matching.
%
For instance, we can calculate the total expected number of constructors |Text|
and |Join| that we will generate by simply expanding the expected number of
generated pattern matching values $simplify\#1$ and $simplify\#2$ into their
corresponding data constructors:
%
{\small
  \begin{align*}
    E^\downarrow\!\!\left[ Text \right]
    = E\left[ Text \right]
    + 2\! \cdot\!\! E\!\left[ simplify\#1 \right]
    + 1\!\! \cdot\!\! E\!\left[ simplify\#2 \right] \\
    E^\downarrow\!\!\left[ Join \right]
    = E\left[ Join \right]
    + 1\!\! \cdot\!\! E\!\left[ simplify\#1 \right]
    + 2\! \cdot\!\! E\!\left[ simplify\#2 \right]
  \end{align*}
}
%
Observe that each time we generate a value satisfying the first pattern matching
of the function |simplify|, we add two |Text| and one |Join| data constructors
to our random value.
%
The case of the second pattern matching of |simplify| follows analogously.
%
Note that the overall prediction cannot be applied if we also generate random
values containing function calls, as we cannot predict the output of an abstract
function without solving the halting problem in the process.

>>>>

Essentially, a branching process is a special kind of Markov process that models
the evolution of a population of \emph{individuals of different kinds} across
discrete time steps known as \emph{generations}.
%
Each kind of individual is expected to produce an average number of offspring of
(possibly) different kinds from one generation to the next one.
%
Mista el at. \cite{DBLP:conf/haskell/MistaRH18} show that branching processes
can be adapted to predict the generation of ADT values by simply considering
each data constructor as a kind of its own.
%
In fact, any ADT value can be seen as a tree where each node represents a root
data constructor and has its sub-expressions as sub-trees---hence note the
similarity with family trees.
%
In this light, each tree level of a random value can be seen as a generation of
individuals in this model.

@@

Vector $E[G_0]$ represents the initial distribuition of constructors that our
generator produces, which simply consists of the generation probability of each
one.
%
The interesting aspect of the prediction mechanism is encoded in the matrix $M$,
known as a the \emph{mean matrix} of this stochastic process.
%
$M$ is a squared matrix with as rows and columns as different data constructors
involved in the generation process.
%
Each element $M_{i,j}$ of this matrix encodes the average number of data
constructors of kind $j$ that gets generated in a given generation, provided that we
generated a constructor of kind $i$ at the previous one.
%
In this sense, this matrix encodes the ``branching'' behavior of our random
generation from one generation to the next one.
%
Each element of the matrix can be automatically calculated by exploiting ADT
definitions. % \cite{DBLP:conf/haskell/MistaRH18}
%
% ---specially the recursive fields---
%
as well as the individual probability of generating each constructor.
%
For instance, the average number of |Text| data constructors that we will
generate provided that we generated a |Join| data constructor on the previous
level is calculated as follows:
%
$$ M_{Join, Text} = 2\cdot p_{Text} $$
%
where $2$ is the number of holes present when generating a partial ADT value
|Join| (i.e., |Join hole hole|) and $p_{Text}$ is the probability of
individually generating the constructor |Text|.
%
This pattern can be used to build the rest of the mean matrix as well.

@@

The new cells are filled as before: we need to consider the amount of holes when
generating partial pattern matching and function calls as well as their
individual probabilities.
%
% Fortunately, it is possible to follow the same pattern as we did on our original
% work, this time considering the amount of recursive pattern variables if it the
% case of a pattern matching value being a father individual, and the amount of
% recursive arguments if we deal with a abstract interface function call as a
% father individual.
%
For instance, if we consider $P_j$ as the second pattern of function |simplify|
and $F_1$ as function |div|, then the marked cell above has the value
$2\cdot p_{div}$, i.e., the amount of holes in the partially generated pattern
|(Join (Join (Text s) hole) hole)|, where |s| is some random string, times the
probability to generate a call to function |div|.
%
The rest of this matrix can be computed analogously.

@@

As another contribution, we found that the whole prediction process can be
factored in terms of two vectors $\beta$ and $\mathcal{P}$, such that $\beta$
represents the number of holes in each partial ADT value that we generate,
whereas $\mathcal{P}$ simply represents the probability of generating that
partial ADT value.% (this probability can be easily calculated based on the
% generation frequencies used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewriten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P})^{n}
\end{align*}
%
% The vector $\beta$ can be obtained by analyzing the shape (namely holes) of each
% source of structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as shown in Fig. \ref{fig:vectors}.
%
We note |simplify#1| and |simplify#2| to the patterns occurring in the first and
second clauses of |simplify|, respectively.
%(Dashed lines indicate the composition of the diffent sub-vectors
%that are calculated separately).

@@

It is possible to provide an overall prediction of the expected number of
constructors when restricting the generation process to only bare data
constructors and pattern matching values.
%
To achieve that, we should stop considering pattern matching values as atomic
values and start seeing them as compositions of several data constructors.
%
In that manner, it is possible to obtain the expected \emph{total} number of
generated data constructors that our generators will produce---regardless if they
are generated on their own, or as part of a pattern matching value.
%
We note this number as $E^\downarrow\![\_]$ and, to calculate it, we only need
to add the expected number of bare constructors that are included within each
pattern matching.
%
For instance, we can calculate the total expected number of constructors |Text|
and |Join| that we will generate by simply expanding the expected number of
generated pattern matching values $simplify\#1$ and $simplify\#2$ into their
corresponding data constructors:
%
{\small
  \begin{align*}
    E^\downarrow\!\!\left[ Text \right]
    = E\left[ Text \right]
    + 2\! \cdot\!\! E\!\left[ simplify\#1 \right]
    + 1\!\! \cdot\!\! E\!\left[ simplify\#2 \right] \\
    E^\downarrow\!\!\left[ Join \right]
    = E\left[ Join \right]
    + 1\!\! \cdot\!\! E\!\left[ simplify\#1 \right]
    + 2\! \cdot\!\! E\!\left[ simplify\#2 \right]
  \end{align*}
}
%
Observe that each time we generate a value satisfying the first pattern matching
of the function |simplify|, we add two |Text| and one |Join| data constructors
to our random value.
%
The case of the second pattern matching of |simplify| follows analogously.
%
Note that the overall prediction cannot be applied if we also generate random
values containing function calls, as we cannot predict the output of an
arbitrary function.
%without solving the halting problem in the process.
